### 2016-02-27 paper/paper.tex
added first version of paper (mostly just copied paper from part 1 and removed parts)

\documentclass[twoside]{article}
\usepackage{aistats2015}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{float}
\usepackage{subfig}
\usepackage{hyperref}
\usepackage{natbib}

\bibpunct{[}{]}{;}{n}{,}{,}

\usetikzlibrary{external} \tikzexternalize[prefix=figures/precompiled/]

    \newlength\figureheight \newlength\figurewidth

\begin{document}

\twocolumn[

\aistatstitle{Learning to Control Kilobots with a Flashlight}
\aistatsauthor{Alexander Hendrich \And Daniel Kauth \And Gregor Gebhardt}
\aistatsaddress{TU Darmstadt \And TU Darmstadt \And TU Darmstadt}]

\begin{abstract}
The recently emerged kilobots provide lots of new applications for swarm
intelligence and collective behavior algorithms. By applying a light following
behavior called \emph{Phototaxis} to the kilobots, they provide an easy method
to move and control large quantities at the same time. By interacting with the
lightsource(s) (switching them on/off or moving a single light source) a human
can control a kilobot swarm to fulfill certain tasks, for example pushing an
object through a maze\cite{kilobotMaze}. In this project we want to learn to
control the light to achieve this behavior.
\end{abstract}

\section{INTRODUCTION}

Our task is to learn to push an object through a maze with a swarm of kilobots
by moving a single light source. Because this task is quite complex and
difficult to learn, we split it up into two separate problems with two separate
policies.

The first problem is to find an optimal path through the labyrinth.
The policy should return a movement direction for each possible object position.
By constantly following these movements, the object should reach the goal
position independent of its starting position.

Because we cannot directly
control the object's movement we need a method to interact with the light
source, so that the kilobots push the object in the desired direction. This will
be the focus of the second problem. The policy should return a new light
position depending on the output of the first policy (the desired movement
direction of the object) and the relative position of the kilobots towards the
object.

\begin{figure}[!htb] \centering
\includegraphics[width=0.8\linewidth]{figures/inspiration.png} \caption{Kilobots
pushing a small object through a maze by human controlled light sources}
\end{figure}

\section{PRELIMINARIES} In the following sections we introduce our used
materials (kilobots\cite{kilobot}) and methods (Actor Critic Relative Entropy
Policy Search)\cite{acreps}.

\subsection{Kilobots} Kilobots\cite{kilobot} are open-source, low cost robots
specially designed to explore collective algorithms and decentralized
cooperation. With previous robots these methods could only be tested in a
simulator or in small quantities of robots because acquiring and controlling
hundreds or even thousands of them is very cost-intensive, time consuming and
complex. To eliminate this issue Harvard University developed Kilobots. Each
robot is made with only 14\$ worth of parts and large collectives can easily be
controlled by a single user.

The low cost however results in limited capabilities of each individual robot.
For example, the kilobot doesn't use complex steering or wheels for locomotion.
Instead it stands on three solid legs and moves by activating two small
vibrating motors placed on each side of the kilobot. This form of movement only
allows for very low movement speeds of up to 1cm per second and can only be
performed on special surfaces (e.g. glass).

To communicate with each other, each kilobot has an infrared transmitter and
receiver on the bottom. Messages are passed by emitting infrared light onto the
reflecting glass and can be received by another kilobot up to 7-10cm away
(radius of each kilobot is 3.3cm). By measuring the received signal strength,
the kilobots are capable of estimating their distances to each other.

The kilobot also possesses an ambient light sensor to determine the light
intensity at its current location. Depending on the current and previously
measured light intensity the robot alternates between left and right turns to
move towards the light source. This behavior is called \emph{Phototaxis} and is
used to control a whole swarm of robots at once. While performing left or right
turns the robot only activates one of its two motors and the robot doesn't move
in a straight line, which further reduces its movement speed.

\begin{figure}[!htb] \centering
    \includegraphics[width=0.9\linewidth]{figures/kilobot.png}
    \caption{Isometric (left) and bottom (right) views of a Kilobot. Some key
    features are: (A) vibration motors, (B) lithium-ion battery, (C) rigid
supporting legs, (D) infrared transmitter/receiver, (E) three-color (RGB) LED,
(F) charging tab, and (G) ambient light sensor. Note the 1 cm line for
scale.\cite{kilobot}} \end{figure}

\subsection{Actor Critic Relative Entropy Policy Search}
\emph{Actor critic relative entropy policy search} (AC-REPS) as introduced by
\cite{acreps} is a method for policy improvement based on \emph{relative entropy
policy search} (REPS) \cite{reps} which allows for model-free learning with
non-parametric continous policies and random exploration. It consists of the
following three steps:

\subsubsection{Estimating the Q-Function}
During the simulation we generate sample in the form $(s_i, a_i, r_i, s'_i)$,
where we start in state $s_i$, take action $a_i$ to get to state $s'_i$ and
get reward $r_i$. We also define a feature-based Q-function as $Q(s, a) =
\phi(s,a)^T \theta$ with parameter vector $\theta$. Details of our state and
action representations and rewards and features we used are discussed in section
TODO.

To estimate $\theta$ we use \emph{Least-squares temporal difference
learning} (LSTD) with $L_{22}-Regularization$\cite{lstdRegularization} as
follows:

With $n$ being the number of sample and $k$ being the number of features
we define the two feature matrices $\Phi$ and $\Phi'$ (size $n\times k$)
$$\Phi = \left(
\begin{array}{c} \phi(s_1,a_1)^T \\ \phi(s_2,a_2)^T \\ \dots \\ \phi(s_n, a_n^T)
\end{array} \right), \quad
\Phi' = \left( \begin{array}{c} \phi(s_1',a_1)^T \\

\phi(s_2',a_2)^T \\ \dots \\ \phi(s_n', a_n^T) \end{array} \right)$$ and the
reward vector
$$R = \left( \begin{array}{c} r(s_1) \\ r(s_2) \\ \dots \\ r(s_n) \end{array} \right).$$

Using the derivations of \cite{lspi} and \cite{lstdRegularization} we can easily
compute $\theta$ as
$$\theta = (X^TX+\beta'I)^{-1}X^Ty$$ with
$$X = C(A+\beta I) \qquad y=Cb$$ $$C = \Phi(\Phi^T\Phi+\beta I)^{-1},$$
where $\beta$ and $\beta'$ are regularization parameters for the projection and
fixed-point step, which make the estimation numerically more stable and improve
noise tolerance.

\subsubsection{Actor Critic REPS}

\subsubsection{Obtaining a new Policy}

\section{LEARNING PROCESS}

\subsection{Object Pushing Policy}

\subsection{Maze solving Policy}

TODO use maze solver or just points list?

\section{FUTURE WORK \& CONCLUSION} Our goal for the second part of this project
is to learn how to control the kilobots, so that they can push a given object
(e.g. flat cylinder) towards the desired direction given by our first policy. To
achieve this goal, the policy first has to move the kilobots behind the object.
Depending on the shape of the object (flat cylinder compared to an ``L''-shaped
object), the optimal positions for the kilobots to push can be different if the
object is rotated.

Our initial approach was to solve the two problems at once, so we already went
through some attempts to model the pushing behavior. Although we could not
achieve satisfactorily results with this approach, we feel confident that our
modeling will return good results once we split up the problem into two distinct
problems. To model the pushing behavior we need to find a good position behind
the object and then push the object with the kilobots. Therefore we placed two
rings of RBFs around the object's position. Because we split up the two policies
we can also increase the number of RBFs we used for the pushing behavior or even
spread them on a grid around the object.

Because we want to apply our learned policies to larger quantities of kilobots,
we will probably have to simulate a large quantity of kilobots during sample
generation. Because the number of simulated kilobots is very crucial to the
performance and time complexity we will have to keep this number quite small.
Maybe we can only simulate a few kilobots or even a single one and let them
represent the center of mass of a whole swarm of kilobots. Otherwise we would
probably have to further improve the simulation performance and extend the
parallel computing capabilities to a larger cluster.

Until now we have learned a policy for moving an object through a maze towards a
goal position. Once the second problem has been solved, we should be able to
easily combine both policies to solve the initial problem of pushing an object
through a maze by controling a swarm of kilobots with a light source. Finally we
want to transfer the resulting policies onto a real world scenario with a self
built maze and real kilobots.

\bibliographystyle{plainnat} \bibliography{bibliography}

\end{document}

### 2016-02-28 learn_single_direction.py
made it possible to restart learning without restarting python (continueLearning=False)

@@ -38,8 +38,9 @@ class MazeLearner:
         self.NUM_NON_KB_DIM = 2
 
         # initial random range for actions
-        aRange = array([[-0.015, 0.015], [-0.015, 0.015]])
-        self.policy = SparseGPPolicy(KilobotKernel(self.NUM_NON_KB_DIM), aRange)
+        self.aRange = array([[-0.015, 0.015], [-0.015, 0.015]])
+        self.policy = SparseGPPolicy(KilobotKernel(self.NUM_NON_KB_DIM),
+                self.aRange)
 
         # kernels used for LSTD
         self.kernelS = KilobotKernel(self.NUM_NON_KB_DIM)
@@ -49,6 +50,7 @@ class MazeLearner:
         self.reps = AC_REPS()
 
         self.it = 0
+        self.S = None
 
     def _sendPolicyModules(self):
         msg = {'message': 'sentPolicyModules',
@@ -235,8 +237,8 @@ class MazeLearner:
         with open(fileName, 'w') as f:
             f.write(pprint.pformat(params, width=1))
 
-    def learn(self, savePrefix, numSampleIt, numLearnIt = 1,
-            startEpsilon = 0.0, epsilonFactor = 1.0):
+    def learn(self, savePrefix, numSampleIt, continueLearning = True,
+            numLearnIt = 1, startEpsilon = 0.0, epsilonFactor = 1.0):
 
         """ sampling """
         self.objectShape = 'quad'
@@ -276,10 +278,20 @@ class MazeLearner:
         self.bwFactorKbGP = factorKb
         self.weightNonKbGP = weightNonKb
 
-
         self.sDim = self.NUM_NON_KB_DIM + 2 * self.numKilobots
-        self.S, self.A, self.R, self.S_ = empty((0, self.sDim)), empty((0, 2)),\
-                empty((0, 1)), empty((0, self.sDim))
+
+        if continueLearning:
+            # dont reset sample or policy
+            if self.S is None:
+                self.S, self.A, self.R, self.S_ = empty((0, self.sDim)),\
+                        empty((0, 2)), empty((0, 1)), empty((0, self.sDim))
+        else:
+            # reset samples, policy and number of iterations
+            self.S, self.A, self.R, self.S_ = empty((0, self.sDim)),\
+                    empty((0, 2)), empty((0, 1)), empty((0, self.sDim))
+            self.policy = SparseGPPolicy(KilobotKernel(self.NUM_NON_KB_DIM),
[…]

### 2016-02-28 paper/paper.tex
added AC-REPS sections

@@
 \usepackage{natbib}
+\usepackage{amsmath}
 
@@
 get reward $r_i$. We also define a feature-based Q-function as $Q(s, a) =
-\phi(s,a)^T \theta$ with parameter vector $\theta$. Details of our state and
-action representations and rewards and features we used are discussed in section
-TODO.
+\phi(s,a)^T \theta$ with feature function $\phi(s,a)$ and parameter vector
+$\theta$. Details of our state and action representations and rewards and
+features we used are discussed in section TODO.
 
@@
 we define the two feature matrices $\Phi$ and $\Phi'$ (size $n\times k$)
-$$\Phi = \left(
-\begin{array}{c} \phi(s_1,a_1)^T \\ \phi(s_2,a_2)^T \\ \dots \\ \phi(s_n, a_n^T)
-\end{array} \right), \quad
-\Phi' = \left( \begin{array}{c} \phi(s_1',a_1)^T \\
-
-\phi(s_2',a_2)^T \\ \dots \\ \phi(s_n', a_n^T) \end{array} \right)$$ and the
-reward vector
-$$R = \left( \begin{array}{c} r(s_1) \\ r(s_2) \\ \dots \\ r(s_n) \end{array} \right).$$
+$$
+\Phi = \left(
+\begin{array}{c}
+    \phi(s_1,a_1)^T \\
+    \phi(s_2,a_2)^T \\
+    \dots \\
+    \phi(s_n,a_n^T)
+\end{array} \right), \;
+\Phi' = \left(
+\begin{array}{c}
+    \mathbf{E}_{\pi(a_1'|s_1')} [\phi(s_1', a_1') | s_1']^T \\
+    \mathbf{E}_{\pi(a_2'|s_2')} [\phi(s_2', a_2') | s_2']^T \\
+    \dots \\
+    \mathbf{E}_{\pi(a_n'|s_n')} [\phi(s_n', a_n') | s_n']^T \\
+\end{array} \right)
+$$
+and the reward vector
+$$
+R = \left(
+\begin{array}{c}
+    r(s_1) \\
+    r(s_2) \\
+    \dots \\
+    r(s_n)
+\end{array} \right).
+$$
+For $\Phi'$ we estimate the expected value
+$$
+\mathbf{E}_{\pi(a_i'|s_i')} [\phi(s_i', a_i') | s_i']
+$$
+by taking the average over $p$ samples $a_i' \sim \pi(a_i'|s_i')$, where $\pi$
+is our current policy.
 
@@
 compute $\theta$ as
-$$\theta = (X^TX+\beta'I)^{-1}X^Ty$$ with
-$$X = C(A+\beta I) \qquad y=Cb$$ $$C = \Phi(\Phi^T\Phi+\beta I)^{-1},$$
+$$
+\theta = (X^TX+\beta'I)^{-1}X^Ty
+$$
+with
+$$X = C(A+\beta I) \qquad y=Cb$$
+$$A = \Phi^T(\phi-\gamma \Phi') \qquad b=\Phi^T R$$
+$$C = \Phi(\Phi^T\Phi+\beta I)^{-1},$$
 where $\beta$ and $\beta'$ are regularization parameters for the projection and
@@
 \subsubsection{Actor Critic REPS}
+In this step we find a new policy $\pi'$ which optimizes the expected Q-value
+as given by our estimated Q-function, but has a limited Kullback-Leibner (KL)
+distance to the old policy $\pi$.
+
+To achieve this we optimize over the joint state action distribution
+$p(s,a) = p(s) \pi'(s, a)$ and require that the estimated state distribution
+$p(s)$ is the same as in the given state sample distribution $\mu(s)$.
+This is implemented by matching feature averages of $p(s)$ and $\mu(s)$ and
+leads to the following optimization problem:
+\begin{align}
+    & \max_p \int p(s, a)Q(s, a)\:ds\:da\:, \\
+    & \text{s.t. } \text{KL}(p(s, a) || \mu(s)\pi(a|s)) \leq \epsilon,\\
+    & \int p(s)\phi(s)\:ds = \hat{\phi}, \int p(s, a)\:ds\:da = 1,
+\end{align}
+where $\hat{\phi} = \int p(s) \phi(s)\:ds$ is the average feature vector of all
+state samples and $\phi(s)$ is another feature function.
+
+This problem can be solved using Lagrangian multipliers and has the solution
+$$
+p(s)\pi'(a|s) \propto \mu(s) \pi(a|s) \exp\left(\frac{Q(s,a) - V(s)}{\eta}\right),
+$$
+where $V(s) = v^T \phi(s)$ and $\eta$ and $v$ are Lagrangian multipliers which
+can be obtained by minimizing the dual function using standard gradient methods.
 
 \subsubsection{Obtaining a new Policy}
+Lastly we need to fit a new policy to our desired probabilities
+$p(s_i, a_i) = p(s_i) \pi'(a_i|s_i)$. This is done by minimizing the KL
+divergence between $\pi'$ and our new policy $\bar{\pi}$ which is equivalent to
+computing a weighted maximum likelihood estimate with sample weights
+$$
+w_i = \exp\left(\frac{Q(s_i, a_i) - V(s_i)}{\eta}\right).
+$$
+
+In this paper we use a weighted Sparse Gaussian Process to obtain our new policy
+$\bar{\pi}$. TODO citation / explain Sparse GP?
 

### 2016-03-07 paper/paper.tex

@@
 
-\usetikzlibrary{external} \tikzexternalize[prefix=figures/precompiled/]
-
-    \newlength\figureheight \newlength\figurewidth
+\newlength\figureheight \newlength\figurewidth % ?
 
@@
 \twocolumn[
-
 \aistatstitle{Learning to Control Kilobots with a Flashlight}
@@
 
-The first problem is to find an optimal path through the labyrinth.
-The policy should return a movement direction for each possible object position.
-By constantly following these movements, the object should reach the goal
-position independent of its starting position.
+The first problem is to find an optimal path through the labyrinth. The policy
+should return a movement direction for each possible object position. By
+constantly following these movements, the object should reach the goal position
+independent of its starting position.
 
-Because we cannot directly
-control the object's movement we need a method to interact with the light
-source, so that the kilobots push the object in the desired direction. This will
-be the focus of the second problem. The policy should return a new light
-position depending on the output of the first policy (the desired movement
-direction of the object) and the relative position of the kilobots towards the
-object.
+Because we cannot directly control the object's movement we need a method to
+interact with the light source, so that the kilobots push the object in the
+desired direction. This will be the focus of the second problem. The policy
+should return a new light position depending on the output of the first policy
+(the desired movement direction of the object) and the relative position of the
+kilobots towards the object.
 
-\begin{figure}[!htb] \centering
-\includegraphics[width=0.8\linewidth]{figures/inspiration.png} \caption{Kilobots
-pushing a small object through a maze by human controlled light sources}
+\begin{figure}[!htb]
+    \centering
+    \includegraphics[width=0.8\linewidth]{figures/inspiration.png}
+    \caption{Kilobots pushing a small object through a maze by human controlled
+        light sources}
 \end{figure}
 
-\section{PRELIMINARIES} In the following sections we introduce our used
-materials (kilobots\cite{kilobot}) and methods (Actor Critic Relative Entropy
-Policy Search)\cite{acreps}.
+\section{PRELIMINARIES}
+In the following sections we introduce our used materials
+(kilobots\cite{kilobot}) and methods (Actor Critic Relative Entropy Policy
+Search)\cite{acreps}.
 
-\subsection{Kilobots} Kilobots\cite{kilobot} are open-source, low cost robots
-specially designed to explore collective algorithms and decentralized
-cooperation. With previous robots these methods could only be tested in a
-simulator or in small quantities of robots because acquiring and controlling
-hundreds or even thousands of them is very cost-intensive, time consuming and
-complex. To eliminate this issue Harvard University developed Kilobots. Each
-robot is made with only 14\$ worth of parts and large collectives can easily be
-controlled by a single user.
+\subsection{Kilobots}
+Kilobots\cite{kilobot} are open-source, low cost robots specially designed to
+explore collective algorithms and decentralized cooperation. With previous
+robots these methods could only be tested in a simulator or in small quantities
+of robots because acquiring and controlling hundreds or even thousands of them
+is very cost-intensive, time consuming and complex. To eliminate this issue
+Harvard University developed Kilobots. Each robot is made with only 14\$ worth
+of parts and large collectives can easily be controlled by a single user.
 
@@
 
-\begin{figure}[!htb] \centering
+\begin{figure}[!htb]
+    \centering
     \includegraphics[width=0.9\linewidth]{figures/kilobot.png}
@@
         features are: (A) vibration motors, (B) lithium-ion battery, (C) rigid
-supporting legs, (D) infrared transmitter/receiver, (E) three-color (RGB) LED,
-(F) charging tab, and (G) ambient light sensor. Note the 1 cm line for
-scale.\cite{kilobot}} \end{figure}
+        supporting legs, (D) infrared transmitter/receiver, (E) three-color
+        (RGB) LED, (F) charging tab, and (G) ambient light sensor. Note the 1 cm
+        line for scale.\cite{kilobot}}
+\end{figure}
 
@@
 $$
-by taking the average over $p$ samples $a_i' \sim \pi(a_i'|s_i')$, where $\pi$
+by taking the average over $m$ samples $a_i' \sim \pi(a_i'|s_i')$, where $\pi$
 is our current policy.
@@
 where $\beta$ and $\beta'$ are regularization parameters for the projection and
-fixed-point step, which make the estimation numerically more stable and improve
+fixed-point step which make the estimation numerically more stable and improve
 noise tolerance.
@@
 leads to the following optimization problem:
-\begin{align}
+\begin{align*}
     & \max_p \int p(s, a)Q(s, a)\:ds\:da\:, \\
@@
     & \int p(s)\phi(s)\:ds = \hat{\phi}, \int p(s, a)\:ds\:da = 1,
-\end{align}
+\end{align*}
 where $\hat{\phi} = \int p(s) \phi(s)\:ds$ is the average feature vector of all
-state samples and $\phi(s)$ is another feature function.
+state samples, $\phi(s)$ is another feature function and $\epsilon$ is the
+maximum allowed distance between the old and new policies.
 
@@
 where $V(s) = v^T \phi(s)$ and $\eta$ and $v$ are Lagrangian multipliers which
-can be obtained by minimizing the dual function using standard gradient methods.
+can be obtained by minimizing the dual function on the samples using standard
+gradient methods.
 
 \subsubsection{Obtaining a new Policy}
-Lastly we need to fit a new policy to our desired probabilities
-$p(s_i, a_i) = p(s_i) \pi'(a_i|s_i)$. This is done by minimizing the KL
-divergence between $\pi'$ and our new policy $\bar{\pi}$ which is equivalent to
-computing a weighted maximum likelihood estimate with sample weights
-$$
-w_i = \exp\left(\frac{Q(s_i, a_i) - V(s_i)}{\eta}\right).
-$$
+The previous step only gives us desired probabilities
+$p(s_i, a_i) = p(s_i) \pi'(a_i|s_i)$ for our samples so in this step we need
+to generalize these by fitting a new policy $\bar{\pi}$.
 
-In this paper we use a weighted Sparse Gaussian Process to obtain our new policy
-$\bar{\pi}$. TODO citation / explain Sparse GP?
+This is done by minimizing the KL divergence between $\pi'$ and our new policy
+$\bar{\pi}$ which is equivalent to computing a weighted maximum likelihood
+estimate with sample weights
+$$ w_i = \exp\left(\frac{Q(s_i, a_i) - V(s_i)}{\eta}\right). $$
+
+In this paper we use a weighted Gaussian Process (GP) to obtain our new policy
+$\bar{\pi}$ by estimating action $a_i$ for state $s_i$ with weight $w_i$. We
+also use the sparse version of the GP to reduce computation time.
+
+This choice of weights will reproduce actions wich lead to a higher reward with
+higher probability but will also choose suboptimal actions with some
+probability, because our new policy can not be too far away in terms of KL
+divergence from the old policy and we start with a random policy. This achieves
+a tradeoff between exploration and exploitation.
+
+TODO citation / explain Sparse GP?
 
@@
 
-\subsection{Object Pushing Policy}
+Our learning process is implemented in Python and uses a simple kilobot
+simulator (see \autoref{sec:simulator}) written by us to generate samples.
 
-\subsection{Maze solving Policy}
+\subsection{Maze Solving Policy}
+This policy tells us where the object should move given the current object
+position and solves the maze (\autoref{fig:simulator}) in this way.
 
+Because this policy is relatively easy to learn and there are specialized maze
+solvers which can easily compute an optimal path through any maze we did not
+learn this policy and instead focused on the second one, which is much more
+interesting. Therefore we manually specified this policy by providing a number
+of waypoints the object should pass and giving the next waypoint as the target
+position. Once we reach a waypoint we move one to next one.
+
+In our experiments this simple policy is sufficient to move different objects
+through a simple maze using our object movement policy. For this reason we will
+only discuss the object moving policy in the following sections.
+
+TODO dont learn first part because already done in first paper?
 TODO use maze solver or just points list?
 
-\section{FUTURE WORK \& CONCLUSION} Our goal for the second part of this project
-is to learn how to control the kilobots, so that they can push a given object
-(e.g. flat cylinder) towards the desired direction given by our first policy. To
-achieve this goal, the policy first has to move the kilobots behind the object.
-Depending on the shape of the object (flat cylinder compared to an ``L''-shaped
-object), the optimal positions for the kilobots to push can be different if the
-object is rotated.
+\subsection{Simulator}
+\label{sec:simulator}
 
-Our initial approach was to solve the two problems at once, so we already went
-through some attempts to model the pushing behavior. Although we could not
-achieve satisfactorily results with this approach, we feel confident that our
-modeling will return good results once we split up the problem into two distinct
-problems. To model the pushing behavior we need to find a good position behind
-the object and then push the object with the kilobots. Therefore we placed two
-rings of RBFs around the object's position. Because we split up the two policies
-we can also increase the number of RBFs we used for the pushing behavior or even
-spread them on a grid around the object.
+\newcommand\nkb{\#\mathit{kb}}
+\newcommand\nep{\#\mathit{ep}}
+\newcommand\nst{\#\mathit{st}}
 
-Because we want to apply our learned policies to larger quantities of kilobots,
-we will probably have to simulate a large quantity of kilobots during sample
-generation. Because the number of simulated kilobots is very crucial to the
-performance and time complexity we will have to keep this number quite small.
-Maybe we can only simulate a few kilobots or even a single one and let them
-represent the center of mass of a whole swarm of kilobots. Otherwise we would
-probably have to further improve the simulation performance and extend the
-parallel computing capabilities to a larger cluster.
+To simulate the kilobots and collect samples for the object movement policy we
+wrote our own simple kilobot simulator (\autoref{fig:simulator}) in Python based
+on the 2D physics engine pybox2d\cite{pybox2d}. This allows us to simulate
+collisions between kilobots and other kilobots, the object and walls.
 
-Until now we have learned a policy for moving an object through a maze towards a
-goal position. Once the second problem has been solved, we should be able to
-easily combine both policies to solve the initial problem of pushing an object
-through a maze by controling a swarm of kilobots with a light source. Finally we
-want to transfer the resulting policies onto a real world scenario with a self
-built maze and real kilobots.
+When requesting new samples from the simulator we specify the current policy
+$\pi$ which will be used to generate these sampels as well as the sampling
+parameters such as the shape of the object, the number of kilobots $\nkb$, the
+number of episodes $\nep$ and the number of steps per episode $\nst$.
 
-\bibliographystyle{plainnat} \bibliography{bibliography}
+Given these parameters the simulator first generates $\nep$ kilobot start
+positions $\mathit{start}_i$ equally spaced in a circle of radius $1.5 \cdot r$
+around the object which starts in the middle of the screen where $r$ is the
+radius or half width of the object depending on the its shape. Then for
+$i \in \{1, \dots, \nep\}$ it does the following:
+
+The object's position is reset to the middle of the screen and its angle is set
+to 0. After that the light position is set to $\mathit{start}_i$ and the
+position of the $\nkb $kilobots is set to $\mathit{start}_i + \mathit{off}_j$,
+where $\mathit{off}_j$ is a small individual offset for kilobot number $j$.
+Then $\nst$ time steps are simulated where in each one an action $a_k$ is chosen
+for the current state $s_k$ according to $\pi$ and applied by adding $a_k$ to
+the current light posiiton. Next the kilobots move directly toward the light and
+lastly we record the current state as $s_k'$, compute the reward $r_k$ and
+save the tuple $(s_k, a_k, r_k, s_k')$ as a sample. We also restrict the
+magnitude of the light and kilobot movements to get a more realistic simulation.
+
+The simulator can also be used to simulate the kilobots pushing an object
+through a maze. In this mode we do not record any samples. Instead we move the
+light according to our object movement policy which gets a target position from
+our maze solving policy. The policies are combined in the following way:
+
+As we will discuss in \autoref{sec:objpolicy} our object movement policy learns
+how to move the light to push the object to the right given the current light
+and kilobot positions relative to the object's position. The maze solving policy
+now gives us a target position $t$ from which we compute the target direction as
+$d = t - o$ where $o$ is the current object position. We mow compute the angle
+$\alpha$ between $d$ and $(1, 0)$ (the vector pointing to the right) and rotate
+the state around $\alpha$. Then we use the object movement policy to get an
+action for our rotated state which we rotate by $-\alpha$ to get the final
+action which we apply just as in the sample generation case.
+
+\begin{figure}[!htb]
+    \centering
+    \includegraphics[width=0.9\linewidth]{figures/simulator_maze.png}
+    \caption{Simulator showing 4 kilobots pushing an object through a maze}
+    \label{fig:simulator}
+\end{figure}
+
+\subsection{Object Movement Policy}
+\label{sec:objpolicy}
+
+% sample generation iteration
+% states
+% features
+
+\section{RESULTS}
+
+\section{FUTURE WORK \& CONCLUSION}
+
+\bibliographystyle{plainnat}
+\bibliography{bibliography}
 

### 2016-03-08 paper/paper.tex
added object movement policy section

@@
 \usepackage{amsmath}
+\usepackage{amssymb}
 
@@
 \subsubsection{Estimating the Q-Function}
+\label{sec:estimateq}
+
 During the simulation we generate sample in the form $(s_i, a_i, r_i, s'_i)$,
@@
 \subsubsection{Actor Critic REPS}
+\label{sec:reps}
+
 In this step we find a new policy $\pi'$ which optimizes the expected Q-value
@@
 \subsubsection{Obtaining a new Policy}
+\label{sec:gp}
+
 The previous step only gives us desired probabilities
@@
 $\bar{\pi}$ by estimating action $a_i$ for state $s_i$ with weight $w_i$. We
-also use the sparse version of the GP to reduce computation time.
+also use the sparse version of the GP as introduced in \cite{sparsegp} to reduce
+computation time.
 
@@
 
-TODO citation / explain Sparse GP?
-
 \section{LEARNING PROCESS}
@@
 Our learning process is implemented in Python and uses a simple kilobot
-simulator (see \autoref{sec:simulator}) written by us to generate samples.
+simulator (see \ref{sec:simulator}) written by us to generate samples.
 
@@
 This policy tells us where the object should move given the current object
-position and solves the maze (\autoref{fig:simulator}) in this way.
+position and solves a maze (\autoref{fig:simulator}) in this way.
 
-Because this policy is relatively easy to learn and there are specialized maze
-solvers which can easily compute an optimal path through any maze we did not
-learn this policy and instead focused on the second one, which is much more
-interesting. Therefore we manually specified this policy by providing a number
-of waypoints the object should pass and giving the next waypoint as the target
-position. Once we reach a waypoint we move one to next one.
+Because this policy is relatively easy to learn and there are specialized path
+planning algorithms like $A^*$ which can easily compute an optimal path through
+any maze we did not learn this policy and instead focused on the second one,
+which is much more interesting. Therefore we manually specified this policy by
+providing a number of waypoints the object should pass and giving the next
+waypoint as the target position. Once we reach a waypoint we move one to next
+one.
 
@@
 through a simple maze using our object movement policy. For this reason we will
-only discuss the object moving policy in the following sections.
-
-TODO dont learn first part because already done in first paper?
-TODO use maze solver or just points list?
+only discuss the object movement policy in the following sections.
 
@@
 radius or half width of the object depending on the its shape. Then for
-$i \in \{1, \dots, \nep\}$ it does the following:
+$i \in \{1, \dots, \nep\}$ we do the following:
 
@@
 to 0. After that the light position is set to $\mathit{start}_i$ and the
-position of the $\nkb $kilobots is set to $\mathit{start}_i + \mathit{off}_j$,
+position of the $\nkb$ kilobots is set to $\mathit{start}_i + \mathit{off}_j$,
 where $\mathit{off}_j$ is a small individual offset for kilobot number $j$.
@@
 for the current state $s_k$ according to $\pi$ and applied by adding $a_k$ to
-the current light posiiton. Next the kilobots move directly toward the light and
-lastly we record the current state as $s_k'$, compute the reward $r_k$ and
-save the tuple $(s_k, a_k, r_k, s_k')$ as a sample. We also restrict the
-magnitude of the light and kilobot movements to get a more realistic simulation.
+the current light positon. Next the kilobots move directly toward the light for
+a short amount time (1 second for all of our simulations) and lastly we record
+the current state as $s_k'$, compute the reward $r_k$ and save the tuple $(s_k,
+a_k, r_k, s_k')$ as a sample. We also restrict the magnitude of the light and
+kilobot movements to get a more realistic simulation.
 
@@
 
-As we will discuss in \autoref{sec:objpolicy} our object movement policy learns
+As we will discuss in \ref{sec:objpolicy} our object movement policy learns
 how to move the light to push the object to the right given the current light
@@
 now gives us a target position $t$ from which we compute the target direction as
-$d = t - o$ where $o$ is the current object position. We mow compute the angle
+$d = t - o$ where $o$ is the current object position. We now compute the angle
 $\alpha$ between $d$ and $(1, 0)$ (the vector pointing to the right) and rotate
@@
 
-% sample generation iteration
-% states
-% features
+For this policy we decided to only learn how to push the object to the right
+instead of an arbitrary direction. This simplifies and speeds up the learning
+process because we need to generate fewer samples and can represent our policy
+using a smaller state subset for the sparse GP and therefore can compute the
+policy update much faster. As discussed in the last paragraph of the previous
+section we can still use this policy to push the object in an arbitrary
+direction by rotation the state and action appropriately.
+
+\subsubsection{On-Policy Learning}
+
+To learn the object movement policy we use on-policy learning. During this
+process we have matrices $S_k, A_k, R_k, S'_k$ with $n$ rows holding the $n$
+current samples which are initially empty and a current policy $\pi_k$. Our
+first policy $\pi_0$ predicts random actions $(a_x, a_y)$, where both $a_x$ and
+$a_y$ are sampled uniformly from $[-0.015, 0.015]$ (1.5 cm).
+
+In each iteration we use the current policy $\pi_k$ to generate $m$ samples
+$(s_i, a_i, r_i, s'_i)$ and add them to the current samples giving matrices
+$S_{k+1}, A_{k+1}, R_{k+1}, S'_{k+1}$ holding $n + m$ samples. To keep the
+computation tractable we then randomly choose $10000$ of the $n + m$ samples
+and replace the current sample matrices by matrices filled with the chosen
+$10000$ samples (if $(n + m) < 10000$ nothing changes).
+
+Next we estimate the Q-function as described in \ref{sec:estimateq},
+compute a sample weighting as in \ref{sec:reps} and lastly fit a new sparse
+GP as discussed in \ref{sec:gp}, which then becomes our new policy $\pi_{k+1}$.
+
+We repeat these steps a fixed number of times and manually inspect the learned
+policy. If it is not satisfactory (it can not solve the maze) we continue the
+learing process.
+
+\subsubsection{States and Actions}
+
+With $n$ kilobots our state tupel is
+$$s = (l_x, l_y, k_{1x}, k_{1y}, \dots, k_{nx}, k_{ny}), $$
+where $(l_x, l_y)$ is the light position and $(k_{ix}, k_{iy})$ is the position
+of the $i-th$ kilobot. All positions are relative to the object's center
+position.
+
+Our actions tuple is
+$$a = (a_x, a_y),$$
+which is the direction the light should move in.
+
+\subsubsection{Rewards}
+
+Our reward is given by
+$$ r = (c_x - p_x) - 0.5 \cdot \mathrm{abs} (c_y - p_y), $$
+where $(c_x, c_y)$ is the object's current position and $(p_x, p_y)$ is the
+object's position from the last time step. Therefore we reward movement to the
+right and punish movement to the left and any movement in $y$ direction.
+
+\subsubsection{Features and Kernels}
+
+For our learing process we need a feature function $\phi(s, a)$ over
+state-action pairs to represent our Q-function, a feature function $\phi(s)$
+over states for REPS and a kernel function $K(s_1, s_2)$ between states for the
+GP.
+
+The feature functions are implemented in terms of kernel functions by
+choosing a random subset of $N$ samples $(s_i, a_i)$ in each iteration and
+computing the feature vectors for a given state $s$ and action $a$ as:
+$$
+\phi(s, a) = \left(
+\begin{array}{c}
+    K'((s_1, a_1), (s, a)) \\
+    K'((s_2, a_2), (s, a)) \\
+    \dots \\
+    K'((s_N, a_N), (s, a))
+\end{array} \right),
+$$
+$$
+\phi(s) = \left(
+\begin{array}{c}
+    K(s_1, s) \\
+    K(s_2, s) \\
+    \dots \\
+    K(s_N, s)
+\end{array} \right),
+$$
+where $K'$ is a kernel function similar to $K$ which compares state-action
+pairs. For our experiments we used $N = 200$.
+
+Let $s$ be a state with $n$ kilobots with positions $(k_{ix}, k_{iy})$
+and $s'$ be a state with $m$ kilobots with positions $(k'_{jx}, k'_{jy})$ and
+let $a$ and $a'$ be the corresponding actions.
+We define $K(s, s')$ and $K'((s, a), (s', a'))$ as follows:
+
+First we compute the distance between the two kilobot configurations as
+\begin{align*}
+d_{kb} =
+  \frac{1}{n^2} &\sum_{i,j \in \{1,\dots,n\}}
+    \bar{K}((k_{ix}, k_{iy}), (k_{jx}, k_{jy})) \\
+- \frac{2}{n \cdot m} &\sum_{\substack{i \in \{1,\dots,n\}\\j \in \{1,\dots,m\}}}
+    \bar{K}((k_{ix}, k_{iy}), (k'_{jx}, k'_{jy})) \\
++ \frac{1}{m^2} &\sum_{i,j \in \{1,\dots,m\}}
+    \bar{K}((k'_{ix}, k'_{iy}), (k'_{jx}, k'_{jy})),
+\end{align*}
+where $\bar{K}$ is a 2 dimensional squared exponential kernel with
+$$
+\bar{K}(k_1, k_2) =
+    \exp \left(-0.5 (k_1 - k_2)^T
+    \mathrm{diag}\left(\frac{1}{\sigma_{kb}^2}\right) (k_1 - k_2)\right)
+$$
+and $\sigma_{kb} \in \mathbb{R}^2$ is the bandwidth for the kilobot dimensions.
+This allows us to compare states with different numbers of kilobots and the
+computed distance is independent of the order of the kilobot positions in the
+state vectors.
+
+After that we compute the distance for the other dimensions (light positions for
+$K$ and light positions and actions for $K'$) as
+$$
+d_{other} = (x_1 - x_2)^T \mathrm{diag}\left(\frac{1}{\sigma_{x}^2}\right) (x_1 - x_2),
+$$
+where $x_1$ and $x_2$ are vectors holding the values for all non-kilobot
+dimensions and $\sigma_x$ is the non-kilobot bandwith.
+
+We combine these distances to get the kernel value
+$$
+\exp(-0.5 (w \cdot d_{other} + (1 - w) \cdot d_{kb})),
+$$
+where $w \in [0,1]$ weighs the importance of the non-kilobot dimensions.
+
+In our experiments we used $w = 0.5$ and recomputed $\sigma_{kb}$ and $\sigma_x$
+in each iteration as the square root of the median of the squared distance
+between all sample values for each dimension separately. We also scaled the
+bandwidths uniformly in each dimension by some manually specified factors.
 

### 2016-03-08 learn_single_direction.py
rewards are now saved

@@ -308,10 +308,15 @@ class MazeLearner:
 
             self.saveParams(os.path.join(savePath, 'params'))
 
+        rewards = []
+
         for i in range(numSampleIt):
+            t = time.time()
+
             # get new samples
             St, At, Rt, S_t = self._getSamples()
             print('sum reward for last samples: {}'.format(Rt.sum()))
+            rewards += [Rt.sum()]
 
             # add samples
             self.S = r_[self.S, St]
@@ -334,8 +339,6 @@ class MazeLearner:
             self.PHI_SA = self.kernelSA.getGramMatrix(SA, self.MuSA)
 
             for j in range(numLearnIt):
-                t = time.time()
-
                 # LSTD to estimate Q function / Q(s,a) = phi(s, a).T * theta
                 self.PHI_SA_ = self._getFeatureExpectation(self.S_, 5, self.MuSA)
                 self.theta = self.lstd.learnLSTD(self.PHI_SA, self.PHI_SA_, self.R)
@@ -351,7 +354,6 @@ class MazeLearner:
 
                 self.it += 1
                 print('finished learning iteration {}'.format(self.it))
-                print('took: {}s'.format(time.time() - t))
 
                 # save results
                 if savePath != '':
@@ -369,10 +371,16 @@ class MazeLearner:
 
             self.epsilon *= self.epsilonFactor
 
+            print('sampling iteration took: {}s'.format(time.time() - t))
+
             gc.collect()
 
+        with open(os.path.join(savePath, 'rewards'), 'w') as f:
+            for r in rewards:
+                f.write('{}\n'.format(r))
+
     def getValueFunctionFigure(self, stepsX = 50, stepsY = 25, N = 4):
-        [X, Y] = meshgrid(linspace(-0.5, 0.5, stepsX), linspace(-0.25, 0.25, stepsY))
+        [X, Y] = meshgrid(linspace(-0.25, 0.25, stepsX), linspace(-0.25, 0.25, stepsY))
         X = X.flatten()
         Y = -Y.flatten()
 
@@ -402,7 +410,7 @@ class MazeLearner:
         return fig
 
     def getPolicyFigure(self, stepsX = 50, stepsY = 25):
-        [X, Y] = meshgrid(linspace(-0.5, 0.5, stepsX), linspace(-0.25, 0.25, stepsY))
+        [X, Y] = meshgrid(linspace(-0.25, 0.25, stepsX), linspace(-0.25, 0.25, stepsY))
         X = X.flatten()
         Y = Y.flatten()
 

### 2016-03-09 learn_single_direction.py
fixed bug where policy parameters are not set

@@ -248,6 +248,23 @@ class MazeLearner:
 
         self.stepsPerSec = 4096
 
+
+        self.sDim = self.NUM_NON_KB_DIM + 2 * self.numKilobots
+
+        if continueLearning:
+            # dont reset sample or policy
+            if self.S is None:
+                self.S, self.A, self.R, self.S_ = empty((0, self.sDim)),\
+                    empty((0, 2)), empty((0, 1)), empty((0, self.sDim))
+        else:
+            # reset samples, policy and number of iterations
+            self.S, self.A, self.R, self.S_ = empty((0, self.sDim)),\
+                    empty((0, 2)), empty((0, 1)), empty((0, self.sDim))
+            self.policy = SparseGPPolicy(KilobotKernel(self.NUM_NON_KB_DIM),
+                    self.aRange)
+            self.it = 0
+
+
         """ LSTD """
         self.lstd.discountFactor = 0.99
 
@@ -277,21 +294,6 @@ class MazeLearner:
         self.bwFactorNonKbGP = factor
         self.bwFactorKbGP = factorKb
         self.weightNonKbGP = weightNonKb
-
-        self.sDim = self.NUM_NON_KB_DIM + 2 * self.numKilobots
-
-        if continueLearning:
-            # dont reset sample or policy
-            if self.S is None:
-                self.S, self.A, self.R, self.S_ = empty((0, self.sDim)),\
-                        empty((0, 2)), empty((0, 1)), empty((0, self.sDim))
-        else:
-            # reset samples, policy and number of iterations
-            self.S, self.A, self.R, self.S_ = empty((0, self.sDim)),\
-                    empty((0, 2)), empty((0, 1)), empty((0, self.sDim))
-            self.policy = SparseGPPolicy(KilobotKernel(self.NUM_NON_KB_DIM),
-                    self.aRange)
-            self.it = 0
 
         self.numLearnIt = numLearnIt
 

### 2016-03-09 paper/paper.tex
wrote results and conclusion sections

@@
 
-\section{FUTURE WORK \& CONCLUSION}
+For our final evaluation we made 20 runs of our learning process with 15
+iterations each generating $25 \cdot 100 = 2500$ samples per iteration. Using
+an Haswell Quad Core with 3.4GHz per core each iteration took about 3 minutes,
+so the total runtime for each policy was about 45 minutes.
+
+The policies start of only being able to push the object when the kilobots start
+to the left of the object. After about 5 iterations they can also handle kilobot
+starting positions above and below the object by first moving the kilobots to
+the left and down and then pushing the object from the left side. After further
+iterations the policies are able to push the object to the right even if the
+kilobots start above and to the right of the object (see
+\autoref{fig:policyEvolution}). With even more iterations the kilobot starting
+positions from which the object can be pushed tend to extend even further, but
+this improvement is inconsistent and the policies may even get worse again.
+Therefore we manually inspected the policies at each iteration and choose a good
+policy (which is typically found in the first about 15 iterations) for solving
+the maze.
+
+\begin{figure*}[!htb]
+	\centering
+	\subfloat[1 iterations]{
+        \def\svgwidth{0.45\textwidth}
+        \input{figures/policy_1.pdf_tex}
+    }
+    \hfill
+	\subfloat[5 iterations]{
+        \def\svgwidth{0.45\textwidth}
+        \input{figures/policy_5.pdf_tex}
+    }
+    \hfill
+	\subfloat[10 iterations]{
+        \def\svgwidth{0.45\textwidth}
+        \input{figures/policy_10.pdf_tex}
+    }
+    \hfill
+	\subfloat[15 iterations]{
+        \def\svgwidth{0.45\textwidth}
+        \input{figures/policy_15.pdf_tex}
+    }
+    \hfill
+	\caption[foo]{Evolution of the learned policy.}
+	\label{fig:policyEvolution}
+\end{figure*}
+
+The value function also shows that kilobot positions left of the object are
+preferrable to positions to the right and you can also somewhat see the value
+function increasing when moving around the object from the right to the left
+(see \autoref{fig:value}).
+
+\begin{figure}[!htb]
+	\centering
+    \def\svgwidth{\columnwidth}
+    \input{figures/value.pdf_tex}
+    \caption{Learned value function after 8 iterations. The object (not shown
+             here) is at the middle of the image.}
+	\label{fig:value}
+\end{figure}
+
+With improving policies the object can be pushed to the right from a larger
+number of kilobot starting positions and therefore the reward tends to increase
+(see \autoref{fig:meanReward}).
+
+\begin{figure}[!htb]
+    \centering
+    \resizebox{\linewidth}{!}{\input{figures/reward.tex}}
+    \caption{Total reward for each iterations averaged over 8 runs.
+             TODO: 20 runs are not complete yet}
+    \label{fig:meanReward}
+\end{figure}
+
+We made the policy plots by getting the mean action for states distributed on a
+grid and plotting it as an arrow indicating the direction the light should move
+in. For this we set the relative light positions to evenly spaced values between
+$-0.25$ and $0.25$ for the $x-$ and $y-$dimension and set the position for all
+kilobots to coincide with the light position.
+
+To compute the Value-function we use the same states and uniformly sample a
+number of actions at random for each state. Then we compute the Q-values for all
+state-action pairs and average them for each state.
+
+To evaluate our policies we used our simulator to push the object through the
+maze as described in \ref{sec:simulator}. We found that our learned policies are
+sufficient to achieve this task and are even able to move the object when the
+kilobots are (almost) completely on the wrong side which requires moving them
+around the object. Moreover we can transfer our policy to more or less kilobots
+and are still able to push the object because of our choice of kernel. We
+learned the policy with 4 kilobots and testes it with 1,2,3,4,6,8 and 10
+kilobots and it was able to push the object through the maze in each case.
+Finally we were also able to push a circel through the maze even though the
+policy was learned on a square.
+
+\section{CONCLUSION \& FUTURE WORK}
+
+In our experiments we achieved our goal to learn a policy to push an object
+using multiple kilobots controlled by a single light and to combine this policy
+with a maze solving policy to push the object through a maze. Our policy is also
+transferable to a different number of kilobots and may be used to push objects of
+different simple symmetric shapes which we tested for a circle.
+
+Although this policy works well in the simulator it is not directly applicable
+to the real kilobots because our simulator is overly simplistic. It moves
+the kilobots directly toward the light instead of using phototaxis and also
+allows for arbitrary light movements where in reality the light would be moved
+by a robot arm which has some constraints on the forces used to move it.
+
+Therefore a future work could simulate the kilobots more realistically and try
+to transfer the learned policy to the real kilobots by building a maze and
+controlling a robot arm holding a flaslight usign the learned policy.
 

### 2016-03-10 paper/paper.tex
added final reward data to the paper
minor changes to the paper

@@
 policy search} (REPS) \cite{reps} which allows for model-free learning with
-non-parametric continous policies and random exploration. It consists of the
+non-parametric continuous policies and random exploration. It consists of the
 following three steps:
@@
 $\theta$. Details of our state and action representations and rewards and
-features we used are discussed in section TODO.
+features we used are discussed in section \ref{sec:objpolicy}.
 
@@
 
-This choice of weights will reproduce actions wich lead to a higher reward with
+This choice of weights will reproduce actions which lead to a higher reward with
 higher probability but will also choose suboptimal actions with some
@@
 divergence from the old policy and we start with a random policy. This achieves
-a tradeoff between exploration and exploitation.
+a trade-off between exploration and exploitation.
 
@@
 which is much more interesting. Therefore we manually specified this policy by
-providing a number of waypoints the object should pass and giving the next
+providing a number of waypoints the object should pass and give the next
 waypoint as the target position. Once we reach a waypoint we move one to next
@@
 When requesting new samples from the simulator we specify the current policy
-$\pi$ which will be used to generate these sampels as well as the sampling
+$\pi$ which will be used to generate these samples as well as the sampling
 parameters such as the shape of the object, the number of kilobots $\nkb$, the
@@
 positions $\mathit{start}_i$ equally spaced in a circle of radius $1.5 \cdot r$
-around the object which starts in the middle of the screen where $r$ is the
-radius or half width of the object depending on the its shape. Then for
+around the object which starts in the middle of the screen, where $r$ is the
+radius or half-width of the object depending on the its shape. Then for
 $i \in \{1, \dots, \nep\}$ we do the following:
@@
 for the current state $s_k$ according to $\pi$ and applied by adding $a_k$ to
-the current light positon. Next the kilobots move directly toward the light for
+the current light position. Next the kilobots move directly toward the light for
 a short amount time (1 second for all of our simulations) and lastly we record
@@
     \includegraphics[width=0.9\linewidth]{figures/simulator_maze.png}
-    \caption{Simulator showing 4 kilobots pushing an object through a maze}
+    \caption{4 kilobots pushing an object through a maze in the simulator.}
     \label{fig:simulator}
@@
 section we can still use this policy to push the object in an arbitrary
-direction by rotation the state and action appropriately.
+direction by rotating the state and action appropriately.
 
@@
 policy. If it is not satisfactory (it can not solve the maze) we continue the
-learing process.
+learning process.
 
@@
 
-With $n$ kilobots our state tupel is
+With $n$ kilobots our state tuple is
 $$s = (l_x, l_y, k_{1x}, k_{1y}, \dots, k_{nx}, k_{ny}), $$
@@
 
-Our actions tuple is
+Our action tuple is
 $$a = (a_x, a_y),$$
@@
 Our reward is given by
-$$ r = (c_x - p_x) - 0.5 \cdot \mathrm{abs} (c_y - p_y), $$
+$$ r = (c_x - p_x) - 0.5 \cdot |c_y - p_y|, $$
 where $(c_x, c_y)$ is the object's current position and $(p_x, p_y)$ is the
@@
 
-For our learing process we need a feature function $\phi(s, a)$ over
+For our learning process we need a feature function $\phi(s, a)$ over
 state-action pairs to represent our Q-function, a feature function $\phi(s)$
@@
 where $x_1$ and $x_2$ are vectors holding the values for all non-kilobot
-dimensions and $\sigma_x$ is the non-kilobot bandwith.
+dimensions and $\sigma_x$ is the non-kilobot bandwidth.
 
@@
 In our experiments we used $w = 0.5$ and recomputed $\sigma_{kb}$ and $\sigma_x$
-in each iteration as the square root of the median of the squared distance
+in each iteration as the square root of the median of the squared distances
 between all sample values for each dimension separately. We also scaled the
@@
 an Haswell Quad Core with 3.4GHz per core each iteration took about 3 minutes,
-so the total runtime for each policy was about 45 minutes.
+so the total run-time for each policy was about 45 minutes.
 
@@
 this improvement is inconsistent and the policies may even get worse again.
-Therefore we manually inspected the policies at each iteration and choose a good
+Therefore we manually inspected the policies at each iteration and chose a good
 policy (which is typically found in the first about 15 iterations) for solving
@@
 	\centering
-	\subfloat[1 iterations]{
+	\subfloat[Policy after 1 iteration]{
         \def\svgwidth{0.45\textwidth}
@@
     \hfill
-	\subfloat[5 iterations]{
+	\subfloat[Policy after 5 iterations]{
         \def\svgwidth{0.45\textwidth}
@@
     \hfill
-	\subfloat[10 iterations]{
+	\subfloat[Policy after 10 iterations]{
         \def\svgwidth{0.45\textwidth}
@@
     \hfill
-	\subfloat[15 iterations]{
+	\subfloat[Policy after 15 iterations]{
         \def\svgwidth{0.45\textwidth}
@@
     \hfill
-	\caption[foo]{Evolution of the learned policy.}
+	\caption[foo]{Evolution of the learned policy}
 	\label{fig:policyEvolution}
@@
 
-The value function also shows that kilobot positions left of the object are
-preferrable to positions to the right and you can also somewhat see the value
-function increasing when moving around the object from the right to the left
-(see \autoref{fig:value}).
+The value function (after 8 iterations) also shows that kilobot positions left
+of the object are preferable to positions to the right. Further it increases
+when moving around the object from the right to the left (see \autoref{fig:value}).
 
@@
     \resizebox{\linewidth}{!}{\input{figures/reward.tex}}
-    \caption{Total reward for each iterations averaged over 8 runs.
-             TODO: 20 runs are not complete yet}
+    \caption{Total reward for each iteration averaged over 20 runs. The shaded
+             area shows twice the standard deviation.}
     \label{fig:meanReward}
@@
 in. For this we set the relative light positions to evenly spaced values between
-$-0.25$ and $0.25$ for the $x-$ and $y-$dimension and set the position for all
+$-0.25$ and $0.25$ for the $x-$ and $y-$dimensions and set the position for all
 kilobots to coincide with the light position.
@@
 sufficient to achieve this task and are even able to move the object when the
-kilobots are (almost) completely on the wrong side which requires moving them
-around the object. Moreover we can transfer our policy to more or less kilobots
-and are still able to push the object because of our choice of kernel. We
-learned the policy with 4 kilobots and testes it with 1,2,3,4,6,8 and 10
-kilobots and it was able to push the object through the maze in each case.
-Finally we were also able to push a circel through the maze even though the
+kilobots are (almost) completely on the wrong side of the object, which requires
+moving them around the object. Moreover we can transfer our policy to more or
+less kilobots and are still able to push the object because of our choice of
+kernel. We learned the policy with 4 kilobots and tested it with 1,2,3,4,6,8 and
+10 kilobots and it was able to push the object through the maze in each case.
+Finally we were also able to push a circle through the maze even though the
 policy was learned on a square.
@@
 to the real kilobots because our simulator is overly simplistic. It moves
-the kilobots directly toward the light instead of using phototaxis and also
-allows for arbitrary light movements where in reality the light would be moved
-by a robot arm which has some constraints on the forces used to move it.
+the kilobots directly toward the light instead of using Phototaxis and also
+allows for arbitrary light movements, where in reality the light would be moved
+by a robot arm which has some constraints on the forces it can apply.
 
@@
 to transfer the learned policy to the real kilobots by building a maze and
-controlling a robot arm holding a flaslight usign the learned policy.
+controlling a robot arm holding a flashlight using the learned policy.
 

### 2016-03-10 paper/paper.tex
some more small paper changes

@@
 higher probability but will also choose suboptimal actions with some
-probability, because our new policy can not be too far away in terms of KL
-divergence from the old policy and we start with a random policy. This achieves
-a trade-off between exploration and exploitation.
+probability. This behavior occurs because our new policy can not be too far away
+in terms of KL divergence from the old policy and we start with a random policy.
+This achieves a trade-off between exploration and exploitation.
 
@@
 
-\newcommand\nkb{\#\mathit{kb}}
-\newcommand\nep{\#\mathit{ep}}
-\newcommand\nst{\#\mathit{st}}
+\newcommand\nkb{n_\mathit{kb}}
+\newcommand\nep{n_\mathit{ep}}
+\newcommand\nst{n_\mathit{st}}
 
-To simulate the kilobots and collect samples for the object movement policy we
-wrote our own simple kilobot simulator (\autoref{fig:simulator}) in Python based
-on the 2D physics engine pybox2d\cite{pybox2d}. This allows us to simulate
-collisions between kilobots and other kilobots, the object and walls.
+We did not use a general purpose robot simulator like V-REP\cite{vrep}, because
+it would take much longer than necessary to generate samples as the kilobots are
+simulated more accurately than needed for our task. Instead we wrote our own
+simple kilobot simulator (\autoref{fig:simulator}) in Python based on the 2D
+physics engine pybox2d\cite{pybox2d}. This allows us to simulate collisions
+between kilobots and other kilobots, the object and walls.
 
@@
 $$
-d_{other} = (x_1 - x_2)^T \mathrm{diag}\left(\frac{1}{\sigma_{x}^2}\right) (x_1 - x_2),
+d_{other} = (x_1 - x_2)^T \mathrm{diag}\left(\frac{1}{\sigma_{other}^2}\right) (x_1 - x_2),
 $$
 where $x_1$ and $x_2$ are vectors holding the values for all non-kilobot
-dimensions and $\sigma_x$ is the non-kilobot bandwidth.
+dimensions and $\sigma_{other}$ is the non-kilobot bandwidth.
 
@@
 
-In our experiments we used $w = 0.5$ and recomputed $\sigma_{kb}$ and $\sigma_x$
-in each iteration as the square root of the median of the squared distances
-between all sample values for each dimension separately. We also scaled the
-bandwidths uniformly in each dimension by some manually specified factors.
+In our experiments we used $w = 0.5$ and recomputed $\sigma_{kb}$ and
+$\sigma_{other}$ in each iteration as the square root of the median of the
+squared distances between all sample values for each dimension separately. We
+also scaled the bandwidths uniformly in each dimension by some manually
+specified factors.
 
@@
 to the left of the object. After about 5 iterations they can also handle kilobot
-starting positions above and below the object by first moving the kilobots to
-the left and down and then pushing the object from the left side. After further
-iterations the policies are able to push the object to the right even if the
-kilobots start above and to the right of the object (see
+starting positions above and below the object by first moving the kilobots
+around to the left side of the object and then pushing the object from the left
+side. After further iterations the policies are able to push the object to the
+right even if the kilobots start above and to the right of the object (see
 \autoref{fig:policyEvolution}). With even more iterations the kilobot starting
@@
 With improving policies the object can be pushed to the right from a larger
-number of kilobot starting positions and therefore the reward tends to increase
+number of kilobot starting positions. Additionally the object can be pushed
+faster and with less vertical movement. Therefore the reward tends to increase
 (see \autoref{fig:meanReward}).

### 2016-03-15 paper/paper.tex
final paper changes

@@
 \twocolumn[
-\aistatstitle{Learning to Control Kilobots with a Flashlight}
+\aistatstitle{Learning to Guide a Swarm of Robots}
 \aistatsauthor{Alexander Hendrich \And Daniel Kauth \And Gregor Gebhardt}
@@
 \begin{abstract}
-The recently emerged kilobots provide lots of new applications for swarm
-intelligence and collective behavior algorithms. By applying a light following
-behavior called \emph{Phototaxis} to the kilobots, they provide an easy method
-to move and control large quantities at the same time. By interacting with the
-lightsource(s) (switching them on/off or moving a single light source) a human
-can control a kilobot swarm to fulfill certain tasks, for example pushing an
-object through a maze\cite{kilobotMaze}. In this project we want to learn to
-control the light to achieve this behavior.
+
+In this work we want to learn a policy which is able to push an object through a
+maze by guiding a swarm of robots. This can be achieved by applying a light
+following behavior called \emph{Phototaxis} to the robots and controlling a
+single light source using the learned policy. We will show that it is useful
+to split up this policy into two parts. The first policy solves the maze while
+the second policy is able to push the object in a single direction. We will
+combine these policies to achieve the original goal in a simulation.
+Finally we will also show that the learned policy can be used to guide a
+different number of kilobots than was used to learn it.
+
 \end{abstract}
@@
 \section{INTRODUCTION}
+
+The recently developed kilobots provide lots of new applications for
+swarm-intelligence and collective behavior algorithms. By applying \emph{Phototaxis}
+to them, they provide an easy method to move and control large quantities at the
+same time. By interacting with a light source a human can control a kilobot swarm
+to fulfill certain tasks, for example pushing an object through a
+maze\cite{kilobotMaze}. In this project we want to learn to control the light
+instead of having to control it manually.
 
@@
     \caption{Kilobots pushing a small object through a maze by human controlled
-        light sources}
+    light sources\cite{kilobotMaze}}
 \end{figure}
@@
 \section{PRELIMINARIES}
-In the following sections we introduce our used materials
+In the following sections we introduce the equipment
 (kilobots\cite{kilobot}) and methods (Actor Critic Relative Entropy Policy
-Search)\cite{acreps}.
+Search)\cite{acreps} we used.
 
@@
 of robots because acquiring and controlling hundreds or even thousands of them
-is very cost-intensive, time consuming and complex. To eliminate this issue
+is very cost-intensive, time consuming and complex. To eliminate this issue,
 Harvard University developed Kilobots. Each robot is made with only 14\$ worth
@@
 vibrating motors placed on each side of the kilobot. This form of movement only
-allows for very low movement speeds of up to 1cm per second and can only be
+allows for very low speeds of up to 1cm per second and can only be
 performed on special surfaces (e.g. glass).
 
-To communicate with each other, each kilobot has an infrared transmitter and
-receiver on the bottom. Messages are passed by emitting infrared light onto the
-reflecting glass and can be received by another kilobot up to 7-10cm away
-(radius of each kilobot is 3.3cm). By measuring the received signal strength,
-the kilobots are capable of estimating their distances to each other.
-
-The kilobot also possesses an ambient light sensor to determine the light
+The kilobot possesses an ambient light sensor to determine the light
 intensity at its current location. Depending on the current and previously
 measured light intensity the robot alternates between left and right turns to
-move towards the light source. This behavior is called \emph{Phototaxis} and is
-used to control a whole swarm of robots at once. While performing left or right
+move towards the light source. This behavior is called
+\emph{Phototaxis} and is used to control a whole swarm of robots at once. For
+more details see \cite{kilobotMaze}. While performing left or right
 turns the robot only activates one of its two motors and the robot doesn't move
@@
 non-parametric continuous policies and random exploration. It consists of the
-following three steps:
+three steps
+\begin{enumerate}
+    \item Estimating the Q-Function
+    \item Actor Critic REPS
+    \item Obtaining a new Policy
+\end{enumerate}
+which will be discussed in he following sections.
 
@@
 
-During the simulation we generate sample in the form $(s_i, a_i, r_i, s'_i)$,
-where we start in state $s_i$, take action $a_i$ to get to state $s'_i$ and
-get reward $r_i$. We also define a feature-based Q-function as $Q(s, a) =
+We assume samples of the form $(s_i, a_i, r_i, s'_i)$,
+where $s_i$ is the current state, $a_i$ is the action taken in this state to get
+to the next state $s'_i$ and $r_i$ is the reward.
+We also have a feature-based Q-function $Q(s, a) =
 \phi(s,a)^T \theta$ with feature function $\phi(s,a)$ and parameter vector
-$\theta$. Details of our state and action representations and rewards and
+$\theta$. Details of the state and action representations and rewards and
 features we used are discussed in section \ref{sec:objpolicy}.
@@
 
-With $n$ being the number of sample and $k$ being the number of features
-we define the two feature matrices $\Phi$ and $\Phi'$ (size $n\times k$)
+The feature matrix for the current state $\Phi$, the expected feature matrix for
+the next state $\Phi'$ and the reward vector $R$ are defined as
+
+$\Phi = \left[\phi(s_1, a_1), \dots, \phi(s_n, a_n)\right]^T,$
 $$
-\Phi = \left(
-\begin{array}{c}
-    \phi(s_1,a_1)^T \\
-    \phi(s_2,a_2)^T \\
-    \dots \\
-    \phi(s_n,a_n^T)
-\end{array} \right), \;
 \Phi' = \left(
@@
 $$
-and the reward vector
-$$
-R = \left(
-\begin{array}{c}
-    r(s_1) \\
-    r(s_2) \\
-    \dots \\
-    r(s_n)
-\end{array} \right).
-$$
-For $\Phi'$ we estimate the expected value
+$R = \left[r(s_1), \dots, r(s_n)\right]^T.$
+
+To compute $\Phi'$ the expected value
 $$
@@
 $$
-by taking the average over $m$ samples $a_i' \sim \pi(a_i'|s_i')$, where $\pi$
-is our current policy.
+is approximated by taking the average over $m$ samples $a_i' \sim
+\pi(a_i'|s_i')$, where $\pi$ is the current policy.
 
-Using the derivations of \cite{lspi} and \cite{lstdRegularization} we can easily
+Using the derivations of \cite{lspi} and \cite{lstdRegularization} one can easily
 compute $\theta$ as
@@
 where $\beta$ and $\beta'$ are regularization parameters for the projection and
-fixed-point step which make the estimation numerically more stable and improve
-noise tolerance.
+fixed-point step, respectively, which make the estimation numerically more
+stable and improve noise tolerance.
 
@@
 In this step we find a new policy $\pi'$ which optimizes the expected Q-value
-as given by our estimated Q-function, but has a limited Kullback-Leibner (KL)
-distance to the old policy $\pi$.
+as given by the estimated Q-function, but has a limited Kullback-Leibner (KL)
+divergence to the old policy $\pi$.
 
 To achieve this we optimize over the joint state action distribution
-$p(s,a) = p(s) \pi'(s, a)$ and require that the estimated state distribution
+$p(s,a) = p(s) \pi'(a|s)$ and require that the estimated state distribution
 $p(s)$ is the same as in the given state sample distribution $\mu(s)$.
@@
 \begin{align*}
-    & \max_p \int p(s, a)Q(s, a)\:ds\:da\:, \\
-    & \text{s.t. } \text{KL}(p(s, a) || \mu(s)\pi(a|s)) \leq \epsilon,\\
-    & \int p(s)\phi(s)\:ds = \hat{\phi}, \int p(s, a)\:ds\:da = 1,
+    \max_p \int p(s, a)Q(s, a)\:ds\:da\:&, \\
+    \text{s.t. } \text{KL}(p(s, a) || \mu(s)\pi(a|s)) &\leq \epsilon,\\
+    \int p(s)\varphi(s)\:ds &= \hat{\phi},\\
+    \int p(s, a)\:ds\:da &= 1,
 \end{align*}
-where $\hat{\phi} = \int p(s) \phi(s)\:ds$ is the average feature vector of all
-state samples, $\phi(s)$ is another feature function and $\epsilon$ is the
-maximum allowed distance between the old and new policies.
+where $\hat{\phi} = \int p(s) \varphi(s)\:ds$ is the average feature vector of all
+state samples, $\varphi(s)$ is another feature function and $\epsilon$ is an
+upper bound on the KL divergence.
 
@@
 $$
-where $V(s) = v^T \phi(s)$ and $\eta$ and $v$ are Lagrangian multipliers which
+where $V(s) = v^T \varphi(s)$ and $\eta$ and $v$ are Lagrangian multipliers which
 can be obtained by minimizing the dual function on the samples using standard
-gradient methods.
+gradient descent.
 
@@
 The previous step only gives us desired probabilities
-$p(s_i, a_i) = p(s_i) \pi'(a_i|s_i)$ for our samples so in this step we need
+$p(s_i, a_i) = p(s_i) \pi'(a_i|s_i)$ for all samples so in this step we need
 to generalize these by fitting a new policy $\bar{\pi}$.
 
-This is done by minimizing the KL divergence between $\pi'$ and our new policy
+This is done by minimizing the KL divergence between $\pi'$ and the new policy
 $\bar{\pi}$ which is equivalent to computing a weighted maximum likelihood
@@
 
-In this paper we use a weighted Gaussian Process (GP) to obtain our new policy
+In this paper we use a weighted Gaussian Process (GP) to obtain the new policy
 $\bar{\pi}$ by estimating action $a_i$ for state $s_i$ with weight $w_i$. We
@@
 higher probability but will also choose suboptimal actions with some
-probability. This behavior occurs because our new policy can not be too far away
+probability. This behavior occurs because the new policy can not be too far away
 in terms of KL divergence from the old policy and we start with a random policy.
@@
 
-Our learning process is implemented in Python and uses a simple kilobot
-simulator (see \ref{sec:simulator}) written by us to generate samples.
+In this section we present the states, actions, features and rewards we used and
+the way we implemented the learning process. Additionally we will describe the
+kilobot simulator that we wrote to generate samples.
 
@@
 providing a number of waypoints the object should pass and give the next
-waypoint as the target position. Once we reach a waypoint we move one to next
+waypoint as the target position. Once we reach a waypoint we move on to the next
 one.
@@
 In our experiments this simple policy is sufficient to move different objects
-through a simple maze using our object movement policy. For this reason we will
+through a simple maze using the object movement policy. For this reason we will
 only discuss the object movement policy in the following sections.
-
-\subsection{Simulator}
-\label{sec:simulator}
-
-\newcommand\nkb{n_\mathit{kb}}
-\newcommand\nep{n_\mathit{ep}}
-\newcommand\nst{n_\mathit{st}}
-
-We did not use a general purpose robot simulator like V-REP\cite{vrep}, because
-it would take much longer than necessary to generate samples as the kilobots are
-simulated more accurately than needed for our task. Instead we wrote our own
-simple kilobot simulator (\autoref{fig:simulator}) in Python based on the 2D
-physics engine pybox2d\cite{pybox2d}. This allows us to simulate collisions
-between kilobots and other kilobots, the object and walls.
-
-When requesting new samples from the simulator we specify the current policy
-$\pi$ which will be used to generate these samples as well as the sampling
-parameters such as the shape of the object, the number of kilobots $\nkb$, the
-number of episodes $\nep$ and the number of steps per episode $\nst$.
-
-Given these parameters the simulator first generates $\nep$ kilobot start
-positions $\mathit{start}_i$ equally spaced in a circle of radius $1.5 \cdot r$
-around the object which starts in the middle of the screen, where $r$ is the
-radius or half-width of the object depending on the its shape. Then for
-$i \in \{1, \dots, \nep\}$ we do the following:
-
-The object's position is reset to the middle of the screen and its angle is set
-to 0. After that the light position is set to $\mathit{start}_i$ and the
-position of the $\nkb$ kilobots is set to $\mathit{start}_i + \mathit{off}_j$,
-where $\mathit{off}_j$ is a small individual offset for kilobot number $j$.
-Then $\nst$ time steps are simulated where in each one an action $a_k$ is chosen
-for the current state $s_k$ according to $\pi$ and applied by adding $a_k$ to
-the current light position. Next the kilobots move directly toward the light for
-a short amount time (1 second for all of our simulations) and lastly we record
-the current state as $s_k'$, compute the reward $r_k$ and save the tuple $(s_k,
-a_k, r_k, s_k')$ as a sample. We also restrict the magnitude of the light and
-kilobot movements to get a more realistic simulation.
-
-The simulator can also be used to simulate the kilobots pushing an object
-through a maze. In this mode we do not record any samples. Instead we move the
-light according to our object movement policy which gets a target position from
-our maze solving policy. The policies are combined in the following way:
-
-As we will discuss in \ref{sec:objpolicy} our object movement policy learns
-how to move the light to push the object to the right given the current light
-and kilobot positions relative to the object's position. The maze solving policy
-now gives us a target position $t$ from which we compute the target direction as
-$d = t - o$ where $o$ is the current object position. We now compute the angle
-$\alpha$ between $d$ and $(1, 0)$ (the vector pointing to the right) and rotate
-the state around $\alpha$. Then we use the object movement policy to get an
-action for our rotated state which we rotate by $-\alpha$ to get the final
-action which we apply just as in the sample generation case.
-
-\begin{figure}[!htb]
-    \centering
-    \includegraphics[width=0.9\linewidth]{figures/simulator_maze.png}
-    \caption{4 kilobots pushing an object through a maze in the simulator.}
-    \label{fig:simulator}
-\end{figure}
 
@@
 
-For this policy we decided to only learn how to push the object to the right
-instead of an arbitrary direction. This simplifies and speeds up the learning
-process because we need to generate fewer samples and can represent our policy
-using a smaller state subset for the sparse GP and therefore can compute the
-policy update much faster. As discussed in the last paragraph of the previous
-section we can still use this policy to push the object in an arbitrary
-direction by rotating the state and action appropriately.
-
-\subsubsection{On-Policy Learning}
-
-To learn the object movement policy we use on-policy learning. During this
-process we have matrices $S_k, A_k, R_k, S'_k$ with $n$ rows holding the $n$
-current samples which are initially empty and a current policy $\pi_k$. Our
-first policy $\pi_0$ predicts random actions $(a_x, a_y)$, where both $a_x$ and
-$a_y$ are sampled uniformly from $[-0.015, 0.015]$ (1.5 cm).
-
-In each iteration we use the current policy $\pi_k$ to generate $m$ samples
-$(s_i, a_i, r_i, s'_i)$ and add them to the current samples giving matrices
-$S_{k+1}, A_{k+1}, R_{k+1}, S'_{k+1}$ holding $n + m$ samples. To keep the
-computation tractable we then randomly choose $10000$ of the $n + m$ samples
-and replace the current sample matrices by matrices filled with the chosen
-$10000$ samples (if $(n + m) < 10000$ nothing changes).
-
-Next we estimate the Q-function as described in \ref{sec:estimateq},
-compute a sample weighting as in \ref{sec:reps} and lastly fit a new sparse
-GP as discussed in \ref{sec:gp}, which then becomes our new policy $\pi_{k+1}$.
-
-We repeat these steps a fixed number of times and manually inspect the learned
-policy. If it is not satisfactory (it can not solve the maze) we continue the
-learning process.
+For the object movement policy we decided to only learn how to push the object
+to the right instead of an arbitrary direction. This simplifies and speeds up
+the learning process because we need to generate fewer samples and can represent
+the policy using a smaller state subset for the sparse GP and therefore can
+compute the policy update much faster. As we will discuss in \ref{sec:simulator}
+we can still use this policy to push the object in an arbitrary direction by
+rotating the state and action appropriately.
 
@@
 
-With $n$ kilobots our state tuple is
+With $n$ kilobots the state tuple is
 $$s = (l_x, l_y, k_{1x}, k_{1y}, \dots, k_{nx}, k_{ny}), $$
@@
 
-Our action tuple is
+The action tuple is given by
 $$a = (a_x, a_y),$$
-which is the direction the light should move in.
+which is the light displacement.
 
@@
 
-Our reward is given by
+The reward is given by
 $$ r = (c_x - p_x) - 0.5 \cdot |c_y - p_y|, $$
 where $(c_x, c_y)$ is the object's current position and $(p_x, p_y)$ is the
-object's position from the last time step. Therefore we reward movement to the
-right and punish movement to the left and any movement in $y$ direction.
+object's position from the last time step. Therefore it rewards movement to the
+right and punishes movement to the left and any movement in $y$ direction.
 
@@
 
-For our learning process we need a feature function $\phi(s, a)$ over
-state-action pairs to represent our Q-function, a feature function $\phi(s)$
+For the learning process we need a feature function $\phi(s, a)$ over
+state-action pairs to represent the Q-function, a feature function $\varphi(s)$
 over states for REPS and a kernel function $K(s_1, s_2)$ between states for the
@@
 choosing a random subset of $N$ samples $(s_i, a_i)$ in each iteration and
-computing the feature vectors for a given state $s$ and action $a$ as:
+computing the feature vectors for a given state $s$ and action $a$ as
 $$
@@
 $$
-\phi(s) = \left(
+\varphi(s) = \left(
 \begin{array}{c}
@@
 let $a$ and $a'$ be the corresponding actions.
-We define $K(s, s')$ and $K'((s, a), (s', a'))$ as follows:
+We define $K(s, s')$ and $K'((s, a), (s', a'))$ as
+$$
+\exp(-0.5 (w \cdot d_{other} + (1 - w) \cdot d_{kb})),
+$$
+where $w \in [0,1]$ weighs the importance of the non-kilobot dimensions and
+$d_{kb}$ and $d_{other}$ are the distances for the kilobot dimensions and all other
+dimensions, respectively. For $K$ the other dimensions are only the light
+positions and for $K'$ they are the light positions and actions.
 
-First we compute the distance between the two kilobot configurations as
+We compute the distance between two kilobot configurations as
 \begin{align*}
@@
 
-After that we compute the distance for the other dimensions (light positions for
-$K$ and light positions and actions for $K'$) as
+The distance for all other dimensions is computed as
 $$
@@
 
-We combine these distances to get the kernel value
-$$
-\exp(-0.5 (w \cdot d_{other} + (1 - w) \cdot d_{kb})),
-$$
-where $w \in [0,1]$ weighs the importance of the non-kilobot dimensions.
+\subsubsection{On-Policy Learning}
 
-In our experiments we used $w = 0.5$ and recomputed $\sigma_{kb}$ and
-$\sigma_{other}$ in each iteration as the square root of the median of the
-squared distances between all sample values for each dimension separately. We
-also scaled the bandwidths uniformly in each dimension by some manually
-specified factors.
+To learn the object movement policy we use on-policy learning. During this
+process we have matrices $S_k, A_k, R_k, S'_k$ with $n$ rows holding the $n$
+current samples which are initially empty and a current policy $\pi_k$. The
+initial policy $\pi_0$ predicts random actions $(a_x, a_y)$, where both $a_x$ and
+$a_y$ are sampled uniformly from $[-0.015, 0.015]$ (1.5 cm).
 
-\section{RESULTS}
+In each iteration we use the current policy $\pi_k$ to generate $m$ samples
+$(s_i, a_i, r_i, s'_i)$.
+We select the new samples $S_{k+1}, A_{k+1}, R_{k+1}, S'_{k+1}$ by randomly
+choosing 10000 samples from the current and newly generated samples to keep the
+computation tractable.
 
-For our final evaluation we made 20 runs of our learning process with 15
+Next we estimate the Q-function as described in \ref{sec:estimateq},
+compute a sample weighting as in \ref{sec:reps} and finally fit a new sparse
+GP as discussed in \ref{sec:gp} to obtain the new policy $\pi_{k+1}$.
+
+We repeat these steps a fixed number of times and manually inspect the learned
+policy. If it is not satisfactory (it can not solve the maze) we continue the
+learning process.
+
+\subsection{Simulator}
+\label{sec:simulator}
+
+\newcommand\nkb{n_\mathit{kb}}
+\newcommand\nep{n_\mathit{ep}}
+\newcommand\nst{n_\mathit{st}}
+
+We did not use a general purpose robot simulator like V-REP\cite{vrep}, because
+it would take much longer than necessary to generate samples as the kilobots are
+simulated more accurately than needed for our task. Instead we implemented our own
+kilobot simulator (\autoref{fig:simulator}) in Python based on the 2D
+physics engine pybox2d\cite{pybox2d}. This allows us to simulate collisions
+between kilobots and other kilobots, the object, and walls respectively.
+
+When requesting new samples from the simulator we specify the current policy
+$\pi$ which will be used to generate these samples as well as the sampling
+parameters such as the shape of the object, the number of kilobots $\nkb$, the
+number of episodes $\nep$ and the number of steps per episode $\nst$.
+
+Given these parameters the simulator first generates $\nep$ light start
+positions $\mathit{start}_i$ equally spaced in a circle of radius $1.5 \cdot r$
+around the object which starts in the middle of the screen, where $r$ is the
+half-width of the object's bounding box. Then for $i \in \{1, \dots, \nep\}$ we
+do the following:
+
+The object's position is reset to the middle of the screen and its angle is set
+to 0. After that the light position $l_k$ is set to
+$\mathit{start}_i$
+and the position of the $\nkb$ kilobots is set to $\mathit{start}_i + \mathit{off}_j$,
+where $\mathit{off}_j$ is a small offset for kilobot $j$.
+Then $\nst$ time steps are simulated where for each an action $a_k$ is chosen
+for the current state $s_k$ according to $\pi$. Subsequently, the light position
+is changed instantly to the new position $l_{k+1} = l_k + a_k$.
+Next the kilobots move directly toward the light for
+a short amount time (1 second for all of our simulations) and finally we record
+the current state as $s_k'$, compute the reward $r_k$ and save the tuple $(s_k,
+a_k, r_k, s_k')$ as a sample. We also restrict the magnitude of the light and
+kilobot movements to get a more realistic simulation.
+
+The simulator can also be used to simulate the kilobots pushing an object
+through a maze. In this mode we do not record any samples. Instead we move the
+light according to the object movement policy which gets a target position from
+the maze solving policy. The policies are combined in the following way:
+
+As discussed in \ref{sec:objpolicy} the object movement policy learns
+how to move the light to push the object right-wards given the current light
+and kilobot positions relative to the object's position. The maze solving policy
+now gives us a target position $t$ from which we compute the target direction as
+$d = t - o$ where $o$ is the current object position. Now, we compute the angle
+$\alpha$ between $d$ and $(1, 0)$ (the vector pointing to the right) and rotate
+the state around $\alpha$. Then we use the object movement policy to get an
+action for the rotated state which we rotate by $-\alpha$ to get the final
+action which we apply just as in the sample generation case.
+
+\begin{figure}[!htb]
+    \centering
+    \includegraphics[width=0.9\linewidth]{figures/simulator_maze.png}
+    \caption{4 kilobots pushing an object through a maze in the simulator.}
+    \label{fig:simulator}
+\end{figure}
+
+\section{EXPERIMENTAL RESULTS}
+
+For the final evaluation we made 20 runs of our learning process with 15
 iterations each generating $25 \cdot 100 = 2500$ samples per iteration. Using
@@
 
-The policies start of only being able to push the object when the kilobots start
+Initially the policies are only able to push the object when the kilobots start
 to the left of the object. After about 5 iterations they can also handle kilobot
@@
 side. After further iterations the policies are able to push the object to the
-right even if the kilobots start above and to the right of the object (see
-\autoref{fig:policyEvolution}). With even more iterations the kilobot starting
-positions from which the object can be pushed tend to extend even further, but
-this improvement is inconsistent and the policies may even get worse again.
-Therefore we manually inspected the policies at each iteration and chose a good
-policy (which is typically found in the first about 15 iterations) for solving
+right even if the kilobots start above and to the right of the object as can be
+seen in \autoref{fig:policyEvolution}. With even more iterations the kilobot
+starting positions from which the object can be pushed tend to extend even
+further, but this improvement is inconsistent and the policies may even get
+worse again.  Therefore we manually inspected the policies at each iteration and
+chose a good
+policy (which is typically found in the first 15 iterations) for solving
 the maze.
@@
     \hfill
-	\caption[foo]{Evolution of the learned policy}
+	\caption[foo]{Evolution of the learned policy. The mean action given by the
+    Gaussian Process is plotted as an arrow for states distributed on a grid.}
 	\label{fig:policyEvolution}
@@
 
-The value function (after 8 iterations) also shows that kilobot positions left
-of the object are preferable to positions to the right. Further it increases
-when moving around the object from the right to the left (see \autoref{fig:value}).
+\autoref{fig:value} also shows that kilobot positions left of the object are
+preferable to positions to the right. Further the value function increases when
+moving around the object from the right to the left.
 
@@
 
-With improving policies the object can be pushed to the right from a larger
+With improving policies the object can be pushed right-wards from a larger
 number of kilobot starting positions. Additionally the object can be pushed
 faster and with less vertical movement. Therefore the reward tends to increase
-(see \autoref{fig:meanReward}).
+as can be seen in \autoref{fig:meanReward}.
 
@@
 
-We made the policy plots by getting the mean action for states distributed on a
+We created the policy plots by getting the mean GP action for states distributed on a
 grid and plotting it as an arrow indicating the direction the light should move
 in. For this we set the relative light positions to evenly spaced values between
-$-0.25$ and $0.25$ for the $x-$ and $y-$dimensions and set the position for all
-kilobots to coincide with the light position.
+$-0.25$ and $0.25$ for the $x-$ and $y-$dimensions, respectively, and set the
+position for all kilobots to coincide with the light position.
 
@@
 number of actions at random for each state. Then we compute the Q-values for all
-state-action pairs and average them for each state.
+state-action pairs and maximize over them for each state.
 
-To evaluate our policies we used our simulator to push the object through the
-maze as described in \ref{sec:simulator}. We found that our learned policies are
+To evaluate the policies we used our simulator to push the object through the
+maze as described in \ref{sec:simulator}. We found that the learned policies are
 sufficient to achieve this task and are even able to move the object when the
 kilobots are (almost) completely on the wrong side of the object, which requires
-moving them around the object. Moreover we can transfer our policy to more or
+moving them around the object. Moreover we can transfer the policy to more or
 less kilobots and are still able to push the object because of our choice of
@@
 using multiple kilobots controlled by a single light and to combine this policy
-with a maze solving policy to push the object through a maze. Our policy is also
+with a maze solving policy to push the object through a maze. The policy is also
 transferable to a different number of kilobots and may be used to push objects of
